---
title: "Online Retail Dataset"
author: "Fintan Cannon"
date: "3/9/2022"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cerulean
    highlight: kate
    self_contained: false
---


  ```{r Import required libraries,echo = FALSE,warning = FALSE,message = FALSE}
library(readr)
library(tidyverse)
library(treemapify)
library(viridis)
library(stringr)
library(rfm)
library(magrittr)
library(snakecase)
library(cowplot)
library(scales)
library(arules)
library(arulesViz)
library(gridExtra)

```

# 1. Introduction

This workbook explores a dataset pertaining to the transactions of a non-store online retailer based in the United Kingdom. 
This workbook will firsty perform basic data exploration of the dataset and implement add value models focusing on CRM management and consumer purchasing patterns.  
  
* Variable Summary:   
  + InvoiceNo: Invoice number. Nominal, a 6-digit integral number uniquely assigned to each transaction. If this code starts with letter 'c', it indicates a cancellation.
  + StockCode: Product (item) code. Nominal, a 5-digit integral number uniquely assigned to each distinct product.
  + Description: Product (item) name. Nominal.
  + Quantity: The quantities of each product (item) per transaction. Numeric.
  + InvoiceDate: Invice Date and time. Numeric, the day and time when each transaction was generated.
  + UnitPrice: Unit price. Numeric, Product price per unit in sterling.
  + CustomerID: Customer number. Nominal, a 5-digit integral number uniquely assigned to each customer.
  + Country: Country name. Nominal, the name of the country where each customer resides.

## 1.2 Goals for this project

Analyze data initially to derive insights that typically add value in an online retail context. 
Features such as stock code, unit price, customer id etc will allow us answer such question as
which SKUs and customers drive revenue the most and when are the busiest times of the year for the retailer. 

The features present will allow us segment customers by performing a Recency, Frequency and Monetary(RFM)
based on their past purchasing behavior. Based on the customer's segmentation the most optimal 
add value action can be executed in relation to each respective customer. 

Finally, a market basket analysis will be performed to identify which SKUs are frequently
purchased together.Knowledge of this will allow for revenue to be maximised by upselling. 


## 1.3 Load data and view variable summaries.


```{r}
retail <- read.csv('C:\\Users\\Finta\\Downloads\\OnlineRetail.csv\\OnlineRetail.csv')
summary(retail)
```


# 2. Preprocessing 

Add additional variables to the data set and standardise the variable names.
The additional variables will allow us analyse the data better and will prove useful when building our respective models. 

```{r}
retail <- retail %>% set_colnames(names(.) %>% to_snake_case())


retail <- retail %>% mutate(
  total_spend = unit_price * quantity, 
  customer_id = as.character(customer_id),
  invoice_hour = as.numeric(format(strptime(invoice_date, "%m/%d/%Y %H:%M"),format = "%H")),
  invoice_date = as.Date(invoice_date, format = '%m/%d/%Y'),
  invoice_month = as.factor(format(invoice_date,"%m")),
  invoice_day = as.factor(format(invoice_date,"%A")), 
  invoice_week = as.factor(format(invoice_date,"%V")),
  refund = str_detect(retail$invoice_no,"^C"),
)


```

* New Variable Notes:
  + refund - Denotes whether observation was canceled transaction and money refunded. Refunds can be identified by having the letter C as the first letter of an observation's invoice number value. 
  + total_spend - Total revenue for each respective transaction. 


## 2.1 Remove obvious errors

Some obvious errors exist such as an extremely large order quantity input which was canceled. 
Some empty transactions exist which have a unit price of zero and customer id value. The value these observations bring is minimal so are removed. 

```{r}

#Remove obvious transaction errors
transaction_errors <- retail %>% filter(total_spend < -5000 | total_spend > 5000) %>%
  pull(total_spend)
`%notin%` <- Negate(`%in%`)
retail <- retail %>% filter(total_spend %notin% transaction_errors)

# Remove 'empty' transactions: Total spend equals 0.00
retail <- retail %>% filter(total_spend != 0.00)

```

## 2.2 Non-transactional records

Using stock code and description variables we can see the rows that contain non-transactional information such as samples. 

```{r}
# Non-transactional stock code & description values

lapply(1:9, function(x){
  j <- retail[which(nchar(retail$stock_code) == x),]
  head(table(j$description))
}) 

```

Counting the most common stock codes according to the number of characters in it.
Stock codes with values of D,M,S represent discounts, manual imput and samples respectively. 

Stock code characters = 1: Discount, Manual input or Sample.  
Stock code characters = 2: 'carriage' primarily pertaining to transactions to Ireland.  
Stock code characters = 3: Dotcom Postage.  
Stock code characters = 4: Postage.  
Stock code characters = 7: Unsaleable, destroyed.  
Stock code characters = 9: Amazon Fee  

```{r}
# Amazon fee
amazon_fee <- retail[which(retail$stock_code == 'AMAZONFEE'),]
amazon_fee_total <- amazon_fee %>% 
  summarise(total_amazon_fee = round(sum(total_spend))) %>%
  pull(total_amazon_fee) %>%
  comma()
# Total Amazon Fee cost
```
Outgoing expense labeled 'AMAZONFEE' totals £`r amazon_fee_total`

```{r}
# Postage
postage_fee <- retail[which(retail$stock_code == 'POST'),]
postage_fee_total <- postage_fee %>%
  summarise(total_postage_fee = round(sum(total_spend))) %>%
  pull(total_postage_fee) %>%
  comma()
```
Revenue labeled 'POST' recorded in primarily non UK transactions totals £`r postage_fee_total`.

```{r}
#Remove chosen values
retail <- retail %>% filter(stock_code %notin% c('S','M'))
retail <- retail[-which(retail$stock_code == 'AMAZONFEE'),]
retail <- retail[-which(retail$stock_code == 'POST'),]
```
Removing transactions pertaining to stock code values for samples, manual input, amazon fees and postage. 

## 2.3 Refunds


```{r,warning = FALSE,message = FALSE}

refund_per_transaction <- round(mean(retail$refund == 'TRUE'),3)
refund_per_transaction
```
Total number of refunds per item bought is `r refund_per_transaction`

```{r,warning = FALSE,message = FALSE}

refund_invoice <-retail %>% filter(refund == 'TRUE') %>% 
  group_by(invoice_no,customer_id) %>%
  summarise(refund_total = sum(total_spend)) %>%
  mutate(refund_total = abs(refund_total)) # Data frame containing refunded invoices only. 

sale_invoice <- retail %>% filter(refund == 'FALSE') %>% 
  group_by(invoice_no,customer_id) %>%
  summarise(refund_total = sum(total_spend)) # Data frame containing sales invoices only. 


# Match refunds with sales based off total invoice amount and customer id
sale_refund_df <- sale_invoice %>% inner_join(refund_invoice,by = c('customer_id','refund_total'))
sale_refund_invoices <- c(sale_refund_df$invoice_no.x,sale_refund_df$invoice_no.y) 
# Combining sales & refund invoice numbers in one vector

retail %>% filter(invoice_no == '537217' | invoice_no == 'C537406') %>% select(-c(invoice_date,invoice_hour,invoice_month, invoice_day,invoice_week))



retail <- retail %>% filter(invoice_no %notin% sale_refund_invoices) 
# Removing invoice numbers pertaining to initial transaction and refund. 


```

Some canceled sales invoices can be matched up with their sales equivalent by matching the customer id and if the absolute value of each invoice is identical. 

## 2.4 Missing values

```{r}
#Missing Values
lapply(retail, function(x){
  missing <- is.na(x)
  length(which(missing == 'TRUE'))
}) #Function counting the number of missing values in each variable
```
customer_id is the only variable with missing values

```{r}
#customer_id missing values
mean(is.na(retail$customer_id))  
```
Proportion of missing values in customer_id variable. 


```{r,warning = FALSE,message = FALSE}
# Total spend comparing existing and missing customer ids
m1 <- retail %>% mutate(missing_cust_id = is.na(customer_id)) %>%
  ggplot(aes(total_spend, fill = missing_cust_id)) +
  geom_histogram() +
  scale_y_continuous(trans = "log10",labels = comma) +
  scale_x_continuous(trans = "log2",labels = comma) +
  xlab("Total Spend") +
  ylab("Count") +
  ggtitle("Total Spend by missing and existing customer id") +
  guides(fill=guide_legend('Missing Customer Id')) 


# Daily transactions comparing exisiting and missing customer ids
m2 <-retail %>% mutate(missing_cust_id = is.na(customer_id)) %>%
  ggplot(aes(invoice_day, fill = missing_cust_id,label = )) +
  geom_histogram(stat = "count") +
  xlab("Day") +
  ylab("Transaction Count") +
  ggtitle("Transactions by day of the week") +
  guides(fill=guide_legend('Missing Customer Id'))

# Country transactions comparing existing and missing customer ids
m3 <-retail %>% mutate(missing_cust_id = is.na(customer_id)) %>%
  ggplot(aes(country, fill = missing_cust_id)) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  guides(fill=guide_legend('Missing Customer Id'))

# Month transactions comparing existing and missing customer ids
m4 <- retail %>% mutate(missing_cust_id = is.na(customer_id)) %>%
  ggplot(aes(invoice_month, fill = missing_cust_id)) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Month") +
  ylab("Transaction Count") +
  ggtitle("Transactions by month") +
  guides(fill=guide_legend('Missing Customer Id'))

# Epoch transactions comparing existing and missing customer ids
m5 <- retail %>% mutate(missing_cust_id = is.na(customer_id)) %>%
  ggplot(aes(invoice_week, fill = missing_cust_id)) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  guides(fill=guide_legend('Missing Customer Id'))

grid.arrange(m1,m2,m4)
```

No discernible reason as to why missing customer id values exist. 


# 3. Exploring Variables 


## 3.1 Invoice_Number
Each individual invoice number represents the purchase(s) made by a customer on a given occasion.

```{r}
unique_invoices <- comma(length(unique(retail$invoice_no)))
# Number of transactions
```
A total of `r unique_invoices` unique invoices contained in this data set



```{r}
# Total spend per invoice
invoice_spend <- retail %>% group_by(invoice_no) %>%
  summarise(total_invoice_amount = sum(total_spend)) %>% 
  arrange(desc(total_invoice_amount))
summary(invoice_spend$total_invoice_amount)

```
Difference between median and mean caused by a number of high revenue generating invoices. 

```{r,warning = FALSE,message = FALSE}
# Total spend per invoice histogram
invoice_hist <- invoice_spend %>%
  ggplot(aes(total_invoice_amount)) +
  geom_histogram(bins = 10) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  geom_vline(xintercept = mean(invoice_spend$total_invoice_amount),        # Add line for mean
             col = "green",
             lwd = 3) +
  geom_vline(xintercept = median(invoice_spend$total_invoice_amount),        # Add line for median
             col = "red",
             lwd = 3) +
  xlab("Total Invoice Amount") +
  ylab("Count") +
  ggtitle("Invoice Amount Histogram")

invoice_hist_log <- invoice_spend %>%
  ggplot(aes(total_invoice_amount)) +
  geom_histogram(bins = 10) +
  scale_x_continuous(trans = 'log2', labels = comma) +
  scale_y_continuous(trans = "log2",labels = comma) +
  geom_vline(xintercept = mean(invoice_spend$total_invoice_amount),       
             col = "green",
             lwd = 3) +                                                     # Add line for mean
  geom_vline(xintercept = median(invoice_spend$total_invoice_amount),        
             col = "red",
             lwd = 3) +                                                     # Add line for median
  xlab("Total Invoice Amount") +
  ylab("Count") +
  ggtitle("Invoice Amount Histogram - Log 2") 

grid.arrange(invoice_hist,invoice_hist_log)


  
```
: Plot 1: Right skewed distribution of the total invoice amounts; a relatively small number of invoices have much larger total amounts than the median invoice amount. 
Plot 2: Scaled distribution of total invoice amount using log2. 

Green Line = Mean: `r round(mean(invoice_spend$total_invoice_amount))`  
Red Line = Median: `r median(invoice_spend$total_invoice_amount)`  

```{r}
invoice_item_quantities <- retail %>% group_by(invoice_no) %>%
 summarise(quantities_bought = sum(quantity)) %>%
  arrange(desc(quantities_bought))
summary(invoice_item_quantities$quantities_bought)

invoice_item_quantities %>% ggplot(aes('',quantities_bought)) +
  geom_boxplot() +
  xlab("") +
  ylab("Quanties per Invoice") +
  ggtitle("Quantity of Items per Invoice") 
```
: Summarized distribution of quantity of items bought per invoice. 


```{r,warning = FALSE,message = FALSE}
# number of items bought per transactions

invoice_quantities <- invoice_item_quantities %>%
  ggplot(aes(quantities_bought)) +
  geom_histogram() +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  geom_vline(xintercept = mean(invoice_item_quantities$quantities_bought),        # Add line for mean
             col = "green",
             lwd = 3) +
  geom_vline(xintercept = median(invoice_item_quantities$quantities_bought),        # Add line for median
             col = "red",
             lwd = 3) +
  xlab("Quantities Bought Per Invoice") +
  ylab("Count") +
  ggtitle("Quantities Per Invoice Histogram") 

invoice_quantities_log <- invoice_item_quantities %>%
  ggplot(aes(quantities_bought)) +
  geom_histogram() +
  scale_x_continuous(trans = 'log2',labels = comma) +
  scale_y_continuous(trans = "log2",labels = comma) +
  geom_vline(xintercept = mean(invoice_item_quantities$quantities_bought),        # Add line for mean
             col = "green",
             lwd = 3) +
  geom_vline(xintercept = median(invoice_item_quantities$quantities_bought),        # Add line for median
             col = "red",
             lwd = 3) +
  xlab("Quantities Bought Per Invoice") +
  ylab("Count") +
  ggtitle("Quantities Per Invoice Histogram - Log2") 

grid.arrange(invoice_quantities,invoice_quantities_log)

```


## 3.2 Stock Code
```{r}
## stock_code
sku_num <- comma(length(unique(retail$stock_code))) #Number of SKUS

```
Number of unique SKUs is `r sku_num`

```{r}

# SKU sales count
sku_sales_count <- retail %>% filter(nchar(retail$stock_code) == 6 |nchar(retail$stock_code) == 7 | nchar(retail$stock_code) == 5) %>%
  count(stock_code)
summary(sku_sales_count$n)

```
Summary of individual SKU sales count  

*Note: Sales count in this context denotes how often an indivual stock code was included in an invoice. 

```{r}
#SKU sales distribution
sku_sales_count  %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = reorder(stock_code,n),n)) +
  geom_bar(stat = 'identity',position="dodge") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  coord_flip() +
  xlab("Stock Code") +
  ylab("SKU Invoice Inclusion") +
  ggtitle("SKU Invoice Inclusion Count") +
  scale_y_continuous(labels = comma)# Conforms strongly to the power rule 
```

SKU purchase distribution conforms strongly to the power rule; a small number of SKUs account for a large percentage of overall SKUs sold. 


```{r}
# SKU sales distribution by quantile - Mosaic graph
sku_sales_count  %>%
  mutate(stock_quant = ntile(rank(n),10)) %>%
  group_by(stock_quant) %>%
  summarise(total_quantile_count = sum(n)) %>%
  mutate(items_sold_proportion = round(total_quantile_count/sum(total_quantile_count),2)) %>%
  ggplot(aes(area = total_quantile_count, fill = stock_quant, label = paste0('Q',stock_quant,'\n\n',items_sold_proportion)))+
  geom_treemap() +
  geom_treemap_text(colour = "Black",
                    place = "centre",
                    size = 20,
                    show.legend = TRUE) +
  theme(legend.position = "none") +
  scale_fill_viridis(option="mako") +
  ggtitle("Quantile Sales Count Proportion")
```
: Each SKU is allocated a quantile based on its respective sales count with the top selling SKUs assigned to Q10.  

The top 10% selling SKUs account for 45% of total SKUs sold.   
The top 20% selling SKUs account for 65% of total SKUs sold.  

```{r,warning = FALSE,message = FALSE}
# Most popular skus
popular_stock_code <- sku_sales_count %>% slice_max(n,n = 20)
popular_stock_code <- popular_stock_code %>% left_join(retail) %>% group_by(stock_code,description,n) %>%
    summarise(total_spend = sum(total_spend)) %>%
  filter(total_spend > 0) %>%
  arrange(desc(n)) %>% 
  rename('total_transaction_number' = n)
popular_stock_code[1:2,]


```

Some stock codes have more than one corresponding description as can be seen above with the highest selling SKU. 



```{r}
popular_stock_code %>%
  arrange(total_transaction_number) %>%
  ggplot(aes(total_transaction_number,fct_inorder(description))) +
  geom_bar(stat = 'identity') +
  xlab("Count of SKUs included in an invoice") +
  ylab("SKU Description") +
  ggtitle("Most Popular SKUs") +
  scale_x_continuous(labels = comma)
```
: Some stock codes have more than one corresponding description as can be seen above with the highest selling SKU. 



```{r,warning = FALSE,message = FALSE}
# Most profitable skus
profitable_stock_code <- popular_stock_code %>% left_join(retail) %>% group_by(stock_code,description,total_transaction_number) %>%
    summarise(total_spend = sum(total_spend)) %>%
  filter(total_spend > 0) %>%
  arrange(desc(total_spend))
profitable_stock_code
```
Highest revenue generating SKUs

```{r}

profitable_stock_code %>%
  arrange(total_spend) %>%
  ggplot(aes(total_spend,fct_inorder(description))) +
  geom_bar(stat = 'identity') +
  xlab("Total Revenue") +
  ylab("SKU Description") +
  ggtitle("Most Profitable SKUs") +
  scale_x_continuous(labels = comma)

```

```{r,warning = FALSE,message = FALSE}
# Non performing SKUs
non_performing_skus <- sku_sales_count %>% filter(n < 4) %>%
  left_join(retail) %>%
  filter(total_spend < 50) %>%
  group_by(stock_code,description) %>%
  summarise(stock_code_revenue = sum(total_spend),number_of_basket_entries = n())
non_performing_skus_nrow <- non_performing_skus %>% nrow()
```
`r non_performing_skus_nrow` SKUs have a total revenue of under £50.

```{r,warning = FALSE,message = FALSE}

# frequently returned stock items 
retail %>% filter(nchar(retail$stock_code) == 6 |nchar(retail$stock_code) == 7) %>%
  filter(refund == 'TRUE') %>%
  count(stock_code) %>%
  arrange(desc(n)) %>%
  top_n(6)

```



## 3.3 Quantity


```{r}
summary(retail$quantity)

```


```{r}
#Distribution of SKU quantities purchased in a transaction
retail %>% 
  filter(quantity > 0) %>%
  ggplot(aes(quantity)) +
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(trans = 'log2',labels = comma) +
  geom_vline(xintercept = mean(retail$quantity),        # Add line for mean
             col = "green",
             lwd = 3) +
    geom_vline(xintercept = median(retail$quantity),        # Add line for median
             col = "red",
             lwd = 3) +
  xlab("Quantities of SKUs bought") +
  ylab("Count") +
  ggtitle("SKU Quantities Bought") +
  scale_y_continuous(labels = comma)
```
: Green Line = Mean: `r round(mean(retail$quantity))`  
Red Line = Median: `r median(retail$quantity)`  

```{r,warning = FALSE,message = FALSE}
quantities_sold_per_invoice <- retail %>% group_by(invoice_no) %>%
  summarise(total_quantities_sold = sum(quantity))
quantities_sold_per_invoice %>% ggplot(aes(total_quantities_sold)) +
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(trans = 'log2',labels = comma) +
  xlab("Quantity Per Invoice") +
  ylab("Count") +
  ggtitle("Quantity Count Per Invoice") +
  scale_y_continuous(labels = comma)
```

## 3.4 Customer ID

```{r}

mean(is.na(retail$customer_id))

```
Total proportion of missing ids is `r mean(is.na(retail$customer_id))`


```{r}
length(unique(retail$customer_id))
```

Total number of unique customer ids is `r length(unique(retail$customer_id))`


```{r,warning = FALSE,message = FALSE}
customer_spend <- retail %>% 
  filter(!is.na(customer_id)) %>%
  group_by(customer_id) %>%
  summarise(total_customer_spend = sum(total_spend)) %>%
  arrange(desc(total_customer_spend)) %>%
  mutate(proportion_of_total_spend = total_customer_spend/sum(total_customer_spend))

customer_spend %>% top_n(10)
```
Highest revenue customers and their proportion of overall revenue. 
```{r}
# Customer spend by quantile
customer_spend %>%
  mutate(quantile = ntile(rank(total_customer_spend),10)) %>%
  group_by(quantile) %>%
  summarise(total_spend = sum(total_customer_spend)) %>%
  mutate(percent_of_sales = total_spend/sum(total_spend),percent_of_sales = scales::percent(percent_of_sales),total_spend = total_spend/1000000) %>%
  ggplot(aes(area = total_spend,fill = quantile,label = paste(percent_of_sales, round(total_spend,digits = 2),paste('Q',quantile), sep = "\n"))) +
  geom_treemap() +
  geom_treemap_text(colour = "Black",
                    place = "centre",
                    size = 20,
                    show.legend = TRUE) +
  theme(legend.position = "none") +
  scale_fill_viridis(option="mako")
```
: Line 1: Percentage of Revenue.  
Line 2: Revenue in millions.  
Line 3: Quantile.  

The top 20% of customers account for around 73% of total revenue. 

## 3.5 Country

```{r}
length(unique(retail$country))
```
Number of countries sold to.

```{r,warning = FALSE,message = FALSE}
customer_country <- retail %>% select(customer_id,country) %>%
  group_by(customer_id) %>%
  summarise(country = unique(country))
table(customer_country$country)
```
Number of customers per country

```{r,warning = FALSE,message = FALSE}
retail %>% group_by(country) %>%
  summarise(spend_per_country = round(sum(total_spend))) %>%
  mutate(prop = round(spend_per_country/sum(spend_per_country),2)) %>% 
  ggplot(aes(area = spend_per_country, fill = country, label = paste(country,comma(spend_per_country),prop,sep = "\n"))) +
  geom_treemap() +
  geom_treemap_text(colour = "Black",
                    place = "centre",
                    size = 20,
                    show.legend = TRUE) +
  theme(legend.position = "none")
```
: Line 1: Country.  
Line 2: Total Revenue.  
Line 3: Proportion of Revenue.  

```{r,warning = FALSE,message = FALSE}
# Non UK Country Spend
retail %>% filter(country != 'United Kingdom') %>%
  group_by(country) %>%
  summarise(spend_per_country = round(sum(total_spend))) %>%
  arrange(spend_per_country) %>%
  ggplot(aes(fct_inorder(country),spend_per_country)) +
  geom_bar(stat = 'identity') +
  scale_y_continuous(trans = 'log2') +
  coord_flip() +
  ylab("Revenue Per Country") +
  xlab("Country") +
  ggtitle("Total Revenue Per Non-UK Country") +
  scale_y_continuous(labels = comma)
```


```{r,warning = FALSE,message = FALSE}
# Customers per country
retail %>% group_by(country) %>%
  summarise(no_of_customers = length(unique(customer_id))) %>%
  arrange(no_of_customers) %>%
  ggplot(aes(fct_inorder(country),no_of_customers)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  ylab("Customers Per Country") +
  xlab("Country") +
  ggtitle("Total Customers Per Country") +
  scale_y_continuous(labels = comma)
```


```{r,warning = FALSE,message = FALSE}
# Customers per country minus UK
retail %>% filter(country != 'United Kingdom') %>%
  group_by(country) %>%
  summarise(no_of_customers = length(unique(customer_id))) %>%
  arrange(no_of_customers) %>%
  ggplot(aes(fct_inorder(country),no_of_customers)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  ylab("Customers Per Country") +
  xlab("Country") +
  ggtitle("Total Customers Per Non-UK Country") +
  scale_y_continuous(labels = comma)
```


```{r,warning = FALSE,message = FALSE}
# Invoices per country
retail %>% group_by(country) %>%
  summarise(invoices_per_country = length(unique(invoice_no))) %>%
  arrange(invoices_per_country) %>%
  ggplot(aes(fct_inorder(country),invoices_per_country)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  ylab("Invoices Per Country") +
  xlab("Country") +
  ggtitle("Total Invoices Per Country") +
  scale_y_continuous(labels = comma)
```


```{r,warning = FALSE,message = FALSE}
# Invoices per country minus UK
retail %>% filter(country != 'United Kingdom') %>%
  group_by(country) %>%
  summarise(invoices_per_country = length(unique(invoice_no))) %>%
  arrange(invoices_per_country) %>%
  ggplot(aes(fct_inorder(country),invoices_per_country)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  ylab("Invoices Per Country") +
  xlab("Country") +
  ggtitle("Total Invoices Per Non-UK Country") +
  scale_y_continuous(labels = comma)
```

## 4. Time Series

```{r,warning = FALSE,message = FALSE}
range(retail$invoice_date)
```
Incomplete calender year for 2011 so dates should range from Dec 2010 to Nov 2011

```{r,warning = FALSE,message = FALSE}
calender_year_sales <- retail %>% filter(between(invoice_date,as.Date('2010-12-01'),as.Date('2011-11-30')))
range(calender_year_sales$invoice_date)
```
Remove dates non pertinent to the above range. 

```{r,warning = FALSE,message = FALSE}
epoch_summary <- calender_year_sales %>% 
  group_by(invoice_week) %>%
  summarise(total_weekly_earnings = sum(total_spend),
            weekly_transactions_no = length(unique(invoice_no)))
epoch_summary_ordered <- epoch_summary[c(48:51,1:47),]

epoch_summary_scaled <- epoch_summary_ordered %>%
  mutate(total_weekly_earnings = scale(total_weekly_earnings),
         weekly_transactions_no = scale(weekly_transactions_no))
head(epoch_summary_ordered)
head(epoch_summary_scaled)

```
Data frames detailing both scaled and total weekly revenue and number of transactions(invoices recorded)


```{r,warning = FALSE,message = FALSE}
epoch_summary_scaled %>%
  gather(factor,value,-invoice_week) %>%
  mutate(factor = as.factor(factor)) %>%
  mutate(factor = recode_factor(factor,total_weekly_earnings = 'Weekly Revenue',weekly_transactions_no = 'Weekly Transactions')) %>%
  ggplot(aes(fct_inorder(invoice_week),value,group= factor,color = factor)) +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  guides(color =guide_legend('')) +
  ylab("Z Score") +
  xlab("Invoice Week") +
  ggtitle("Transaction and Revenue Weekly Distribution") +
  scale_y_continuous(labels = comma)
```
: Weekly revenue heavily correlated with the number of transactions.  
Week 37 to 41 sees a divergence from this correlation. Possible that some wholesalers place large revenue generating orders during this window. 


```{r,warning = FALSE,message = FALSE}
epoch_summary_earnings <- epoch_summary_ordered %>% 
  select(invoice_week,total_weekly_earnings) %>%
  gather(total_weekly_earnings,value,-invoice_week) %>%
  ggplot(aes(fct_inorder(invoice_week),value,group = total_weekly_earnings)) +
  geom_point() +
  geom_line() +
  geom_smooth() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ylab("Revenue") +
  xlab("Invoice Week") +
  ggtitle("Total Weekly Revenue") +
  scale_y_continuous(labels = comma)
epoch_summary_earnings
```
: Invoice revenue per week. 

```{r,warning = FALSE,message = FALSE}
# Total sales by month
monthly_earnings <- calender_year_sales %>% group_by(invoice_month) %>%
  summarise(earnings_by_month = sum(total_spend))
monthly_earnings <- monthly_earnings[c(12,1:11),]
monthly_earnings %>%
  ggplot(aes(fct_inorder(invoice_month),earnings_by_month,group = 1)) +
  geom_line() +
  ylab("Revenue") +
  xlab("Invoice Month") +
  ggtitle("Total Montly Revenue") +
  scale_y_continuous(labels = comma)
  
```

# 5. RFM Analysis

```{r,warning = FALSE,message = FALSE}
arb_date <- max(retail$invoice_date) + 1

rfm_retail <- retail %>% filter(!is.na(customer_id)) %>%
  group_by(invoice_no,customer_id,invoice_date) %>%
  summarise(total_spend = sum(total_spend))
```

Prepare data frame for RFM analysis by removing rows with missing customer id values.

```{r,warning = FALSE,message = FALSE}
#RFM Table
rfm_data  <- rfm_table_order(
  data = rfm_retail,
  customer_id = customer_id,
  order_date = invoice_date,
  revenue = total_spend,
  analysis_date = arb_date,
  recency_bins = 5,
  frequency_bins = 5,
  monetary_bins = 5,
)
rfm_data[1:6,]
```
Assign RFM score to each customer id. 
```{r,warning = FALSE,message = FALSE}
# Assign segments names
segment_names <- c("Champions", "Loyal Customers", "Potential Loyalist",
                   "New Customers", "Promising", "Need Attention", "About To Sleep",
                   "At Risk", "Can't Lose Them", "Hibernating", "Lost")

# Assign classification rules for segments
recency_lower   <- c(4, 2, 3, 4, 3, 3, 2, 1, 1, 2, 1)
recency_upper   <- c(5, 4, 5, 5, 4, 4, 3, 2, 1, 3, 1)
frequency_lower <- c(4, 3, 1, 1, 1, 3, 1, 2, 4, 2, 1)
frequency_upper <- c(5, 4, 3, 1, 1, 4, 2, 5, 5, 3, 1)
monetary_lower  <- c(4, 4, 1, 1, 1, 3, 1, 2, 4, 2, 1)
monetary_upper  <- c(5, 5, 3, 1, 1, 4, 2, 5, 5, 3, 1)

segments <- rfm_segment(rfm_data, segment_names, recency_lower, recency_upper,
                        frequency_lower, frequency_upper, monetary_lower, monetary_upper)
```

```{r,warning = FALSE,message = FALSE}
# Combine customer segmentation with main data frame
combined_rfm_df <- retail %>% left_join(segments) %>% 
  select(customer_id,total_spend,segment,rfm_score,recency_score,frequency_score,monetary_score,stock_code)
combined_rfm_df <- combined_rfm_df %>% group_by(customer_id) %>%
  summarise(total_spend = sum(total_spend),segment = unique(segment),rfm_score = unique(rfm_score),
            recency_score = unique(recency_score),frequency_score = unique(frequency_score), monetary_score = unique(monetary_score))
combined_rfm_df
```

```{r,warning = FALSE,message = FALSE}
# Count of customers per section
combined_rfm_df %>% filter(!is.na(customer_id)) %>%
  group_by(segment) %>%
  summarise(count = n(), spend = round(sum(total_spend)))
```

```{r,warning = FALSE,message = FALSE}
segment_count <- combined_rfm_df %>% filter(!is.na(customer_id)) %>%
  group_by(segment) %>%
  summarise(count = n(), spend = sum(total_spend)) %>%
    arrange(count) %>%
  ggplot(aes(count,fct_inorder(segment))) +
  geom_bar(stat = 'identity') +
  ylab("Segment") +
  xlab("Count") +
  ggtitle("Total Segment Count") +
  scale_x_continuous(labels = comma)

segment_spend <- combined_rfm_df %>% filter(!is.na(customer_id)) %>%
  group_by(segment) %>%
  summarise(count = n(), spend = sum(total_spend)) %>%
    arrange(spend) %>%
  ggplot(aes(spend,fct_inorder(segment))) +
  geom_bar(stat = 'identity') +
  ylab("Segment") +
  xlab("Revenue") +
  ggtitle("Total Segment Revenue") +
  scale_x_continuous(labels = comma)
  
grid.arrange(segment_count,segment_spend) 
  
```

# 6. Market Basket Analysis

## 6.1 Feature: Description

Model to identify what products are likely to be purchased together and be used to develop or optimize  cross selling initiatives on retail sites. 

```{r,warning = FALSE,message = FALSE}
# Preprocess data for basket analysis - Description
retail_basket <- retail %>% filter(refund == 'FALSE') %>%
  select(invoice_no,description)
retail_basket <- retail_basket %>% group_by(invoice_no) %>%
  mutate(id = cur_group_id()) %>% 
  ungroup() %>%
  select(id, description) # Assign basket variable as a key. 
```

```{r,warning = FALSE,message = FALSE}
# Convert data to class transaction
retail_list <- split(retail$description,retail_basket$id) # Group items according the basket they belong to
retail_trans <- as(retail_list,'transactions') #Convert list to transactions
retail_rules <- apriori(retail_trans, parameter=list(supp=0.007, conf=0.4)) # Create rule object


# Visualise rules
plot(retail_rules)
```
: Support: The percentage of transactions that contain all of the items in an itemset.   
Confidence: The probability that a transaction that contains the items on the left hand side of the rule also contains the item on the right hand side.  
Lift: The increase in likelihood of the right hand side item of the rule being purchased given the left hand side item(s) of the rule are purchased.   
```{r,warning = FALSE,message = FALSE}
# Inspect rules
inspect(head(sort(retail_rules,by='lift'),35))
```

```{r,warning = FALSE,message = FALSE}
retail_hi_lift <- head(sort(retail_rules, by = 'lift'),50)
plot(retail_hi_lift,method = 'graph',col = 'blue',cex = 1)
```

```{r,warning = FALSE,message = FALSE}
itemFrequencyPlot(retail_trans,topN=30,type = 'absolut',horiz = TRUE)
```
: Most frequently purchased items.    
## 6.2 Feature: Stock Code

Model is applied in the same way except the variable now used in stock code. 
```{r,warning = FALSE,message = FALSE}
# Preprocess data for basket analysis - Stock Code
retail_basket_stock <- retail %>% filter(refund == 'FALSE') %>%
  select(invoice_no,stock_code)
retail_basket_stock <- retail_basket_stock %>% group_by(invoice_no) %>%
  mutate(id = cur_group_id()) %>% 
  ungroup() %>%
  select(id, stock_code) # Assign basket variable as a key. 
summary(retail_basket_stock)
```

```{r,warning = FALSE,message = FALSE}
# Convert data to class transaction
retail_list_stock <- split(retail$stock_code,retail_basket_stock$id) # Group items according the basket they belong to
retail_trans_stock <- as(retail_list_stock,'transactions') #Convert list to transactions
retail_rules_stock <- apriori(retail_trans_stock, parameter=list(supp=0.007, conf=0.4)) # Create rule object

```

```{r,warning = FALSE,message = FALSE}
# Visualise rules
plot(retail_rules_stock)
```

```{r,warning = FALSE,message = FALSE}
# Inspect rules
inspect(head(sort(retail_rules_stock,by='lift'),35))
```

```{r,warning = FALSE,message = FALSE}
retail_hi_lift_stock <- head(sort(retail_rules_stock, by = 'lift'),50)
plot(retail_hi_lift_stock,method = 'graph',col = 'blue',cex = 1)
```

```{r,warning = FALSE,message = FALSE}
itemFrequencyPlot(retail_trans_stock,topN=30,type = 'absolut',horiz = TRUE)
```




